---
title: "Reanalysing Hsu & Bishop"
author: "DVM Bishop"
date: "15/05/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(plyr)
require(tidyverse)
require(stargazer)
require(beanplot)
require(rddtools)
```
## Background
The usual way of looking for implicit learning in serial-reaction-time (SRT) task has been to compare performance when there is a repeated pattern in the stimuli (learning phase) with performance when the pattern is broken (random phase). Typically this has been done by comparing RT for the last block in a pattern phase with the first block in the random phase: if the pattern was learned, we should see 'rebound' when it is broken. This approach was adopted in a paper by Hsu and Bishop (2014), which used a 'Green Monster' task developed by Tomblin et al (2007). The child was presented with four squares in a line and had to press a key to correspond to the square in which a green monster appeared. After an initial random phase of 100 trials, there were 200 trials where the repeated sequence 2 4 1 3 4 2 1 4 3 1 was used, and then a further 100 random trials. (Note: this sequence was wrongly described in the published paper: this is the correct sequence).

In a later paper, Kuppuraj et al (2018) used a task designed to overcome some issues with previous research. In particular, a square array of four pictures was used and the task was to select the named picture. Using this method, it was possible to interleave different types of sequential dependency to compare learning for short sequences that varied in the statistical structure of dependencies. We also adopted a new analytic approach that used all of the data, regression discontinuity. This uses a test for the difference in slope between a pattern phase and a random phase. Although regression discontinuity is usually applied to group data, the method of comparing two slopes can be used for data from a single case, and it gives a t-value for each participant and condition that can be evaluated to determine whether learning occurred (i.e. whether the difference in the two slopes is statistically reliable). 

In this paper, the main goal is apply the regression discontinuity approach to the original Hsu and Bishop dataset, to compare its sensitivity relative to the original method of measuring implicit learning.

The definitive version of the dataset for Hsu and Bishop (2014), alldata_GM.xlsx, is available on Open Science Framework here: https://osf.io/396ai/wiki/home/. A csv version with just the worksheet showing responses and RTs is saved here.



```{r readdata, echo=FALSE}
## Read SRT data
#  worksheet has response number in initial block, and then RTs below, with subjects in columns. NaN used for RT for inaccurate responses.

filename <- 'Alldata_GM.csv'
all_dat <- data.frame(read.csv(filename,stringsAsFactors = FALSE))
rt_dat<-all_dat[401:800,1:97]
stim_dat<-all_dat[1:400,1:97]
#stargazer(rt_dat,type='text') #check appearance of data - 
#comment this out to see stats for individual subjects
```



```{r getmeds, echo=FALSE}
## Reformat to get medians
#Reformat file to get out medians for blocks of 20 trials.
#Should replicate Figure 2 in Hsu and Bishop.
mysubs <- colnames(rt_dat[2:97])
med.dat <- data.frame(matrix(NA,nrow=length(mysubs),ncol=23))
colnames(med.dat) <- c('ID','group','R1','R2','R3','R4','R5',
                        'P1','P2','P3','P4','P5',
                        'P6','P7','P8','P9','P10',
                        'R1a','R2a','R3a','R4a','R5a','errorrate')
med.dat$ID <- mysubs
med.dat$group <- substr(mysubs,1,1) #peel off first character
med.dat$groupnum<-1 #to avoid nightmare of trying to turn factor level to number later on
w<-which(med.dat$group=='L')
med.dat$groupnum[w]<-2
w<-which(med.dat$group=='S')
med.dat$groupnum[w]<-3

for (j in 1:length(mysubs)){ 
  readcol <- which(colnames(rt_dat)==mysubs[j])
for (i in 1:20){
  b1 <- 1+(i-1)*20
  b2 <- (i*20)
  med.dat[j,i+2] <- median(as.integer(rt_dat[b1:b2,readcol]),na.rm=TRUE)
 }
}
```
```{r missing, echo=FALSE}
## Get error rate
#Identify any cases to exclude (overall accuracy < 80%).
#NB. Dev Sci paper reports excluding 5 lang match and 2 SLI but I get 5 lang match and 1 SLI

for (i in 1:length(mysubs)){
  med.dat$errorrate[i] <- length(which(is.na(rt_dat[,(i+1)])))/400
}
#recode group for those with >19% errors, so not included later
w <- which(med.dat$errorrate>.19)
med.dat$group[w] <- 'X' 
med.dat$groupnum[w]<-4
med.dat$group <-as.factor(med.dat$group)
```

## Data summary for three groups
This plot shows the data processed as in the original Hsu and Bishop paper: i.e. medians for correct responses extracted for blocks of 20 trials.

```{r getmeans, echo=FALSE}
## Get means by group
#And plot lines
range1 <- which(med.dat$group=='C')
range2 <- which(med.dat$group=='L')
range3 <- which(med.dat$group=='S')


meantable <- rbind(colMeans(med.dat[range1,3:22]),
                   colMeans(med.dat[range2,3:22]),
                   colMeans(med.dat[range3,3:22]))
# get the range for y axis
yrange <- c(200,1600)
xrange <- c(1,20)

# set up the plot
plot(xrange, yrange, type="n", xlab="block",
   ylab="RT" )
colors <- rainbow(3)
linetype <- c(1:3)
plotchar <- seq(19,21,1)

# add lines
jrange=matrix(c(1,5,6,15,16,20),byrow=TRUE, nrow=3)
for (i in 1:3) {
  for (j in 1:3){
    thisrange <-jrange[j,1]:jrange[j,2]
  mygroup <- meantable[i,thisrange]
  lines(thisrange,mygroup, type="b", lwd=1.5,
    lty=linetype[i], col=colors[i], pch=plotchar[i])
}
}
# add a title and subtitle
title("SRT task: Medians by block of 20")

# add a legend
legend(3, 600, c('Age-match','Language match','SLI'), cex=0.8, col=colors,
   pch=plotchar, lty=linetype, title="Group")
text(2,1400,'Random')
text(7,1400,'Pattern')
text(17,1400,'Random')
```



```{r writefile, echo=FALSE}
## Write med.dat to file for OSF - this step not now run
#write.csv(med.dat,'GreenMonster_MedRT.csv')
```

## Original subtraction method
We compute the difference between the last datapoint for the pattern blocks and the first datapoint for the random blocks, and compare this for the 3 groups. NB. We use the acronym DLD for 'developmental language disorder', which has superseded SLI (for 'specific language impairment') -see Bishop et al (2017).

```{r subtract, echo=FALSE}
levels(med.dat$group) = c('Age-control','Lang-control','DLD','Excluded')

med.dat$diffc <- med.dat[,18]-med.dat[,17]
beanplot(diffc ~ group, data = med.dat, col = "lightgray", border = "grey", 
        ylab='Diff RT from last Pattern to first Random')
```

## Regression discontinuity approach
We compare the original approach with the method of Kuppuraj and Bishop.
The package rddtools (Stigler & Quast, 2015) is used to compare the slopes of two portions of data. We do this for the pattern and second random sequence. The t-value for this comparison can be used as an index of learning. This table shows the N children from each group who learned (p < .05, one-tailed - i.e. negative t-values ignored).

```{r regcoeff, echo=FALSE}
## Regression discontinuity analysis for individual participants
nsubs <- nrow(med.dat)
med.dat$reg_t <-NA #initialise columns to hold results
med.dat$reg_p <-NA
for (i in 1:nsubs){
  myrange<-8:22 #cols with Pattern and Rand2 trials
  mycut=18 #start col for pattern
  rdd.temp<-rdd_data(y=as.numeric(med.dat[i,myrange]),x=myrange,cutpoint=mycut)   # using the RDDtools package

  reg_para <- rdd_reg_lm(rdd_object = rdd.temp, order = 1) #this is just linear: for higher order can increase
  st<-summary(reg_para)[[4]]
  myt<-st[2,3]#t-value corresponding to difference in slope for two phases
  myp<-summary(reg_para)$coefficients[2,4]#sig of slope diff for the two phases
 med.dat$reg_t[i]<-myt
 med.dat$reg_p[i]<-myp
}

#divide according to whether reg_p is < .05
med.dat$psig <- 0
w<-unique(which(med.dat$reg_p<.1),which(med.dat$reg_t>0)) #onesided test - ignore neg t values, so use p < .1
med.dat$psig[w]<-1
learntab<-table(med.dat$group,med.dat$psig)
colnames(learntab)<-c('Not learned','Learned')
learntab
```

Now we look at the distribution of scores for t-values. This appears to give clearer differentiation between groups than the original analysis.



```{r plottvalues, echo=FALSE}
beanplot(reg_t ~ group, data = med.dat, col = "lightgray", border = "grey", 
        ylab='Diff RT from last Pattern to first Random')
mycor<-round(cor(med.dat$diffc,med.dat$reg_t),3) #correlation between difference and regression methods, rounded to 3 decimals

```

### Compare discontinuity regression results with subtraction results
The relationship between the two measures is shown here.

```{r correlmethods,echo=FALSE}

plot(med.dat$diffc,med.dat$reg_t,xlab='Original RT difference',ylab='t-value from regression',col=med.dat$group)
legend(-400, 4, legend=levels(med.dat$group),pch=1,
       col=1:4,  cex=0.8)
text(400,-2,paste0('r = ',mycor))
```

### Anova with diffc and with reg_t for 3 groups
```{r anova, echo=FALSE}
aov.dat <-  filter(med.dat, group !='Excluded')
myaov <- aov(diffc~group,data=aov.dat)
myaov2 <- aov(reg_t~group,data=aov.dat)
print('Original analysis on difference score')
summary(myaov)

print('New analysis on t-value for regression discontinuity')
summary(myaov2)
```




### Compute effect sizes (d) for different methods
Pairwise comparisons with age-control for (i) DLD/age-con and (ii) lang-con/age-con
```{r effsize, echo=FALSE}

dtab<-ddply(aov.dat,~group,summarise,N=length(diffc),Mean.diff=mean(diffc),sd=sd(diffc))
dtab
sd1<-dtab[1,4]
sd2<-dtab[2,4]
sd3<-dtab[3,4]
N1<-dtab[1,2]
N2<-dtab[2,2]
N3<-dtab[3,2]
poolsd13<-sqrt(((N1-1)*sd1^2+(N3-1)*sd3^2)/(N1+N3-2))
teff1 <- round((dtab[1,3]-dtab[3,3])/poolsd13,2)


poolsd12<-sqrt(((N1-1)*sd1^2+(N2-1)*sd2^2)/(N1+N2-2))
teff2 <- round((dtab[1,3]-dtab[2,3])/poolsd12,2)

#Now repeat with t-values

dtab<-ddply(aov.dat,~group,summarise,N=length(reg_t),Mean.t=mean(reg_t),sd=sd(reg_t))
dtab
sd1<-dtab[1,4]
sd2<-dtab[2,4]
sd3<-dtab[3,4]
N1<-dtab[1,2]
N2<-dtab[2,2]
N3<-dtab[3,2]
poolsd13<-sqrt(((N1-1)*sd1^2+(N3-1)*sd3^2)/(N1+N3-2))
teff3 <- round((dtab[1,3]-dtab[3,3])/poolsd13,2)


poolsd12<-sqrt(((N1-1)*sd1^2+(N2-1)*sd2^2)/(N1+N2-2))
teff4 <- round((dtab[1,3]-dtab[2,3])/poolsd12,2)

```

The regression method yields larger effect sizes, as would be expected given the ANOVA results.
With the original method, for the DLD vs age-matched controls, Cohen's d is `r teff1`, and for age-matched controls vs language-matched controls it is `r teff2`. 

With the regression discontinuity method, for the DLD vs age-matched controls, Cohen's d is `r teff3`, and for age-matched controls vs language-matched controls it is `r teff4`. 

## Conclusion
The regression discontinuity approach appears more sensitive to detecting the rebound effect in a serial RT task than the traditional approach of subtracting the RT from the final block of the pattern phase from the first block of the subsequent random phase.

## Additional analysis of item-specific effects
The sequence used in the Green Monster task was 10 items long, raising concerns that the individual responses were not of equal frequency. In the sequence 2 4 1 3 4 2 1 4 3 1, numbers 2 and 3 occur twice, whereas 1 and 4 occur thrice. We were interested in the possibility that this difference in frequency of occurrence might influence the RTs. In fact, when the responses were subdivided into two subgroups, corresponding to stimuli 2 and 3 vs stimuli 1 and 4, it was apparent from observation that responses to the latter stimuli were indeed more rapid. However, this difference was clearly present in the first random block, when all stimuli were equally frequent. This suggests that the effect is not due to stimulus frequency, but rather to perceptual crowding: the four stimuli were arranged in a linear row, with 1 and 4 at the ends. They would therefore be more discriminable than stimuli 2 and 3, which had another box on either side.

```{r reformat, echo=FALSE}
mysubs <- colnames(rt_dat[2:97])
med.dat2 <- data.frame(matrix(NA,nrow=length(mysubs),ncol=43))
colnames(med.dat2) <- c('ID','group','R1.1','R2.1','R3.1','R4.1','R5.1',
                        'P1.1','P2.1','P3.1','P4.1','P5.1',
                        'P6.1','P7.1','P8.1','P9.1','P10.1',
                        'R1a.1','R2a.1','R3a.1','R4a.1','R5a.1',
                       'R1.2','R2.2','R3.2','R4.2','R5.2',
                        'P1.2','P2.2','P3.2','P4.2','P5.2',
                        'P6.2','P7.2','P8.2','P9.2','P10.2',
                        'R1a.2','R2a.2','R3a.2','R4a.2','R5a.2','errorrate')
med.dat2$ID <- mysubs
med.dat2$group <- substr(mysubs,1,1) #peel off first character


for (j in 1:length(mysubs)){ 
  readcol <- which(colnames(rt_dat)==mysubs[j])
  writecol <- 2 #first col will be 3
       for (k in 1:2){
         thisk <- c(2,3)
         if(k>1) {thisk <- c(1,4)}
for (i in 1:20){
  b1 <- 1+(i-1)*20
  b2 <- (i*20)
    thisseq <- stim_dat[b1:b2,readcol]
     thisrt <- rt_dat[b1:b2,readcol]

       w <- which(thisseq%in%thisk)
       writecol<-writecol+1
  med.dat2[j,writecol] <- median(as.integer(thisrt[w]),na.rm=TRUE)
 }
}
}
```

```{r makemeans, echo=FALSE}
meantable <- rbind(colMeans(med.dat2[range1,3:22],na.rm=TRUE),
                   colMeans(med.dat2[range1,23:42],na.rm=TRUE),
                   colMeans(med.dat2[range2,3:22],na.rm=TRUE),
                   colMeans(med.dat2[range2,23:42],na.rm=TRUE),
                   colMeans(med.dat2[range3,3:22],na.rm=TRUE),
                   colMeans(med.dat2[range3,23:42],na.rm=TRUE))
```

```{r plotsubsets,echo=FALSE}
# get the range for y axis
yrange <- c(200,1600)
xrange <- c(1,20)

# set up the plot
plot(xrange, yrange, type="n", xlab="block",
   ylab="RT" )
colors <- rainbow(3)
linetype <- c(1:3)
plotchar <- c(1,3)

# add lines
jrange=matrix(c(1,5,6,15,16,20),byrow=TRUE, nrow=3)
thisrow <- 0

for (i in 1:3) { #3 groups
  for (k in 1:2){ #2 stimulus types (1/4 or 2/3)
  thisrow <- thisrow+1
  for (j in 1:3){ #ranges for 3 types of trial
    thisrange <-jrange[j,1]:jrange[j,2]
  mygroup <- meantable[thisrow,thisrange]
  lines(thisrange,mygroup, type="b", lwd=1.5,
    lty=linetype[i], col=colors[i], pch=plotchar[k])
}
}
}
# add a title and subtitle
title("SRT task")

# add a legend
legend(3, 600, c('Age-match','Language match','SLI'), cex=0.8, col=colors,
   pch=plotchar, lty=linetype, title="Group")
text(2,1400,'Random')
text(7,1400,'Pattern')
text(17,1400,'Random')
text(15,400,'Crosses are stimuli 1/4;\n circles are 2/3')

```

This provides an opportunity to look at the regression discontinuity t-value separately for each subset. This will be based on half as much data so will be noisier, but could give an indication of split half reliability of the method.

The first table shows agreement in terms of categorical classification in relation to p < .05: this is not very good. 

```{r splithalf,echo=FALSE}
med.dat2$reg_t1 <-NA #initialise columns to hold results
med.dat2$reg_t2 <-NA
med.dat2$reg_p1 <-NA #initialise columns to hold results
med.dat2$reg_p2 <-NA
med.dat2$psig1 <-0
med.dat2$psig2 <-0
thiscol<-which(names(med.dat2)=='reg_t1') #find first col to write to
 for (j in 1:2){ #do both patterns one after the other
for (i in 1:nsubs){
 
  myrange<-8:22 #cols with Pattern and Rand2 trials
  mycut=18 #start col for pattern
  if (j==2){
    myrange=28:42
    mycut=38
  }
  rdd.temp<-rdd_data(y=as.numeric(med.dat2[i,myrange]),x=myrange,cutpoint=mycut)   # using the RDDtools package

  reg_para <- rdd_reg_lm(rdd_object = rdd.temp, order = 1) #this is just linear: for higher order can increase
  st<-summary(reg_para)[[4]]
  myt<-st[2,3]#t-value corresponding to difference in slope for two phases
  myp<-summary(reg_para)$coefficients[2,4]#sig of slope diff for the two phases
  colmat<-matrix(c(44, 45,46,47,48,49),nrow=2)
 med.dat2[i,colmat[j,1]]<-myt
 med.dat2[i,colmat[j,2]]<-myp
 if(myt>0 && myp<.1){med.dat2[i,colmat[j,3]]<-1}
}
 }
rbit<-cor(med.dat2$reg_t1,med.dat2$reg_t2)

agreetab<-table(med.dat2$psig1,med.dat2$psig2)

colnames(agreetab)<-c('nolearn','learn')
rownames(agreetab)<-c('nolearn','learn')
agreetab
```

The correlation between the t-values for the two item sets is moderate, r = `r round(rbit,3)`.

```{r plotsplithalf,echo=FALSE}
plot(med.dat2$reg_t1,med.dat2$reg_t2,xlab='t-value for 2 & 3',ylab='t-value  for 1 & 4',col=med.dat$groupnum,main='Regression discontinuity t-values for 2 item subsets')
legend(1.8,-1.5, legend=levels(med.dat$group),pch=1,
       col=1:4,  cex=0.8)
text(-4,2,paste0('r = ',round(rbit,3)))
```

## References
Bishop, D. V. M., Snowling, M. J., Thompson, P. A., Greenhalgh, T., & CATALISE Consortium. (2017). Phase 2 of CATALISE: a multinational and multidisciplinary Delphi consensus study of problems with language development: Terminology. Journal of Child Psychology and Psychiatry, 58(10), 1068-1080. doi:10.1111/jcpp.12721

Hsu, H. J., & Bishop, D. V. M. (2014). Sequence- specific procedural learning deficits in children with specific language impairment. Developmental Science, 17(3), 352-365. doi:10.1111/desc.12125.

Kuppuraj, S., Duta, M., Thompson, P. A., & Bishop, D. V. M. (2018). Online incidental statistical learning of audiovisual word sequences in adults - a registered report Royal Society Open Science, 5(2), 171678. doi:10.1098/rsos.171678
 
Stigler, M & Quast, B. (2015). rddtools: Toolbox for Regression Discontinuity Design ('RDD'). R package version 0.4.0. https://CRAN.R-project.org/package=rddtools

Tomblin, J. B., Mainela-Arnold, E., & Zhang, X. (2007). Procedural learning in adolescents with and without specific language impairment. Language Learning and Development,, 3, 269-293. 


##Session information
```{r sessinfo}
sessionInfo()
```
